---
title: "Accessing SpaMTP Experiment Data"
author: "Tianyao Lu"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Accessing SpaMTP Experiment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`SpaMTPData` makes the experiment data used in SpaMTP workflows discoverable
through ExperimentHub. The package itself remains small and users download only
the resources required for a particular workflow.

The submission API uses camelCase function names. Install SpaMTPData >= 0.99.4,
SpaMTPdb >= 0.99.3 and SpaMTP >= 0.99.4 together to replace the historical API.
Package versions and the version of a published dataset are independent.

```{r setup}
library(SpaMTPData)
resources <- spaMTPDataResources(version = "latest")
resources[c("resource", "version", "category", "r_data_class", "requires_conversion")]
spaMTPDataVersion()
```

# A native example without downloads

`spaMTPExampleData()` constructs a small, deterministic synthetic
SpatialExperiment. It is not a biological dataset and does not load Seurat.
Use it to check the container interface or run the three-package workflow
demonstrated in SpaMTP's `SpaMTP_Companion_Workflow` vignette.

```{r native-example}
spe <- spaMTPExampleData()
SummarizedExperiment::assayNames(spe)
SummarizedExperiment::rowData(spe)$mz
SummarizedExperiment::colData(spe)$region
SpatialExperiment::spatialCoords(spe)
SingleCellExperiment::altExpNames(spe)

core <- spe[, spe$region == "core"]
stopifnot(identical(colnames(SingleCellExperiment::altExp(core, "transcriptome")),
  colnames(core)))
```

# Published native datasets

The maintainer recipe `inst/scripts/prepare-native-resources.R` extracts the
seven archival Seurat datasets into native release `1.1.0`, now published in
[Zenodo record 22733262](https://zenodo.org/records/22733262).
It retains counts/data, feature and pixel annotations, image-centroid
coordinates and paired transcriptomes. The resulting SpatialExperiment RDS
can be read directly without Seurat or any further conversion:

```{r published-native, eval=FALSE}
striatum <- spaMTPData("mouse_brain_dhb_striatum", version = "1.1.0")
SummarizedExperiment::assayNames(striatum)
SpatialExperiment::spatialCoords(striatum)
SingleCellExperiment::altExpNames(striatum) # SPT
```

This downloading example is not evaluated during a package build. The
accompanying native manifest records source and output hashes. Optical rasters,
polygons, graphs, reductions and Seurat-specific weight objects are omitted;
the retained `data` layer is not reinterpreted as log-normalized expression.
The simulated datasets retain their explicit synthetic status.

The resource API selects `1.1.0` by default. Native files have new filenames
and checksums and are not substituted under an old archive's identity.
Use `version = "1.0.0"` to request the original immutable archives explicitly.
Hub ingestion is separate from publication and package metadata registration.

The native files are published in the same Zenodo version family as
SpaMTPdb's annotation resources. SpaMTPdb coordinates file publication;
SpaMTPData retains experimental resource ownership, documentation and reading.
There is no runtime dependency between the two data packages. Registration of
a published native release preserves the 1.0.0 history and carries forward all
18 resources, including eleven unchanged files at their original locations.

# Loading a published native dataset

Inspect the mouse brain/DHB record without downloading the object:

```{r inspect-resource}
striatum_metadata <- spaMTPData(
  "mouse_brain_dhb_striatum",
  metadata = TRUE
)
striatum_metadata[c("resource", "version", "r_data_class", "bytes")]
stopifnot(striatum_metadata$r_data_class == "SpatialExperiment",
  !striatum_metadata$requires_conversion)
```

Omit `metadata = TRUE` to retrieve and cache the native object. ExperimentHub
is preferred; an immutable source-URL fallback is available before ingestion:

```r
striatum <- spaMTPData("mouse_brain_dhb_striatum")
```

This default file is a SpatialExperiment and does not need Seurat. Historical
metadata can be inspected without downloading the old Seurat object:

```{r inspect-historical-resource}
historical <- spaMTPData("mouse_brain_dhb_striatum",
  version = "1.0.0", metadata = TRUE)
historical[c("resource", "version", "r_data_class", "requires_conversion")]
```

Historical entries marked `requires_conversion = TRUE` need explicit
conversion at the SpaMTP workflow boundary. Downloading an old file does not
update its scientific processing. Its class and checksum remain unchanged.

The same interface can use a previously downloaded local file:

```r
options(SpaMTPData.resource_dir = "/path/to/SpaMTPData-files")
striatum <- spaMTPData("mouse_brain_dhb_striatum", offline = TRUE)
```

Local files and caches are verified against the recorded size and MD5.
`offline = TRUE` also uses previously verified download-cache entries; it never
queries ExperimentHub or Zenodo. A missing or invalid file is an error, not a
reason to silently use another version. `verify = FALSE` permits intentional
local development fixtures but does not bypass object-class validation.

The explicit resource accessor follows the same interface:

```{r explicit-accessor}
spaMTPDataResource(
  "mouse_brain_dhb_striatum",
  metadata = TRUE
)
```

# Session information

```{r session-info}
sessionInfo()
```
